Notes

### 2022-08-26 thesis/chapters/background-related-work.tex
Revision up to chapter 4 complete

 The most common scalar-on-function linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the monograph by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^p\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is a random error term independent from \(X\) with \(\E [\epsilon]=0\) and \(\E[X(t)\epsilon]=0\) almost everwhere; and the functional slope parameter \(\beta=\beta(\cdot)\) is assumed to be a member of the infinite-dimensional space \(L^2[0, 1]\). A careful rearrangement of~\eqref{eq:l2-linear-model} shows that the model is equivalently expressed as \(\Delta=\mathcal K\beta\), where \(\Delta(\cdot)\) is the cross-covariance function of \(X\) and \(Y\), and \(\mathcal K\) is the covariance operator of the process \(X\) (to be defined in Section~\ref{sec:rkhs}). This expression is a continuous analog of the normal equations that arise in finite-dimensional regression settings, and since the operator \(\mathcal K\) is non-invertible in general, questions of existence and unicity of \(\beta\) need further study (see \citet{cardot2011functional} for more insights).

@@

A common inference strategy is to expand both \(X\) and \(\beta\) on a certain data-driven orthonormal basis of \(L^2[0,1]\), say \(\{\phi_j\}\), up to a certain value \(p\in\N\):
\[
  X(t)=\sum_{j=1}^p x_j \phi_j(t), \quad \beta(t) = \sum_{j=1}^p \beta_j \phi_j(t).
\]
In this way, model~\eqref{eq:l2-linear-model} simplifies to
\[
Y = \alpha_0 + \sum_{j=1}^p \beta_j x_j,
\]
which can then be solved in the usual manner. For example, when the chosen basis is composed of eigenfunctions of the covariance operator \(\mathcal K\), this method is known as Functional Principal Component Regression (FPCR). Another class of methods focus on selecting an \textit{a priori} basis for \(\beta\) (e.g. a spline basis) and solving a penalized least squares problem. In particular, for a truncated basis representation of the form \(\beta(t)=\phi(t)'b\), with \(\phi(t)=(\phi_1(t), \dots, \phi_p(t))'\) and \(b\in\R^p\), a typical minimization problem would be
\[
\argmin_{(\alpha_0, b)\in\R\times\R^p} \ \left\{ \sum_{i=1}^n \left( y_i - \alpha_0 - \int_0^1 x_i(t)[\phi(t)'b]\, dt\right)^2 + \lambda \Omega(b) \right\},
\]
where \(\Omega\) is some penalty function and \(\lambda>0\) is a regularization parameter.

Turning our attention to logistic regression, a similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \Prob(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][Sec.~3.2]{berrendero2018functional}.

@@

\begin{outcomment}
  Rellenar?
\end{outcomment}

@@

In this section we present a brief exposition of some basic concepts regarding reproducing kernel Hilbert spaces (RKHS's). Although there are quite a few ways of introducing these spaces, we adopt a probabilistic point of view that will set the stage for the development of the subsequent theory. For a more detailed account, a good reference is the book by~\citet{berlinet2004reproducing}.

@@

\begin{definition}
  We say that a function of two variables \(K: \mathcal T \times \mathcal T \to \R\) is positive semidefinite\footnote{Sometimes in the literature these functions are known simply as positive definite.} if
  \[
    \sum_{i,j=1}^p a_ia_jK(t_i, t_j) \geq 0
  \]
  for any \(p\in\N\), any \((a_1,\dots, a_p)\in \R^p\) and any \((t_1,\dots, t_p)\in\mathcal T^p\). Note that this is equivalent to saying that the matrix \((K(t_i, t_j))_{i,j}\) is positive semidefinite for any choice of \(p\in\N\) and \((t_1,\dots, t_p)\in\mathcal T^p\).
\end{definition}

@@

\begin{proposition} The following properties hold for the space \(\Hcal(K)\):
  \begin{enumerate}
    \item The evaluation operator \(\delta_t\) is bounded for all \(t\in[0,1]\).
    \item Norm convergence implies pointwise convergence, and if \(K\) is continuous, it also implies uniform convergence.
    \item If \(K\) is \(m\)-times continuously differentiable, then so is every function \(f\in\Hcal(K)\). In particular, if \(K\) is continuous, every function \(f\in\Hcal(K)\) is continuous.
  \end{enumerate}
\end{proposition}
\begin{proof}
  To prove \((i)\), simply note that for any \(t\in[0,1]\) and \(f\in\Hcal(K)\) the Cauchy-Schwarz inequality and the reproducing property tell us that
  \[
  |\delta_t(f)| = |f(t)| = |\dotprod{f}{K(t, \cdot)}_K| \leq \|K(t, \cdot)\|_K \|f\|_K = \sqrt{K(t, t)}\|f\|_K,
  \]
  and consequently \(\|\delta_t\|\leq \sqrt{K(t, t)}\). To see \((ii)\), observe that for any \(t\in[0,1]\) we have
  \[
    |f_n(t) - f(t)| = |\delta_t(f_n - f)| \leq \|\delta_t\|\|f_n - f\|_K,
  \]
  and we have just shown above that the evaluation operator is bounded. Moreover, if \(K\) is continuous on its (compact) domain, \(\|\delta_t\| \leq \sup_s \sqrt{K(s, s)}=M<\infty\) for all \(t\), so the convergence is indeed uniform. Finally, we prove the second statement in \((iii)\), and refer the reader to \citet[][Th 2.6]{saitoh2016theory} for a complete proof. If \(K\) is continuous, then it is clear that every function in \(\Hcal_0(K)\) is continuous. Let us now fix an arbitrary \(f\in \Hcal(K)\). By definition, there exists a sequence \(\{f_n\}\subset \Hcal_0(K)\) of continuous functions such that \(\|f_n - f\|_K \to 0\). But then \(f_n \to f\) uniformly by \((ii)\), and therefore \(f\) is continuous as the uniform limit of continuous functions. Note that this statement remains valid even when the underlying domain is not compact.
\end{proof}

@@

\begin{example}[Finite-dimensional RKHS]
  Setting aside our stochastic process framework for a moment, we can conceive examples outside of \(\mathcal T =[0,1]\). Consider a vector-valued random variable \(X = (X_1, \dots, X_p)\) with non-singular covariance matrix \(\Sigma\). In this case the index set would be \(\mathcal T_p = \{1,\dots,p\}\), so identifying functions on \(\mathcal T_p\) with points in the \(p\)-dimensional Euclidean space, we have \(\Hcal(\Sigma)=\R^p\), with inner product \(\dotprod{x}{y}_\Sigma=x\Sigma^{-1}y\).
\end{example}

@@

 Note that, as expected, the covariance function of \(X\) plays a crucial role in characterizing the associated RKHS. An integral operator closely related to the kernel is the so-called covariance operator, namely
\[
\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f \in L^2[0, 1],
\]
which is self-adjoint and compact when \(K\) is continuous \citep[e.g.][Th.~4.6.2]{hsing2015theoretical}, so for the remaining of this work we will indeed suppose that \(K\) is continuous. As it turns out, the covariance function admits a spectral decomposition in terms of the eigenvalues and eigenfunctions of this integral operator, a sort of continuous generalization of the eigendecomposition of a symmetric positive semidefinite matrix.

\begin{theorem}[Mercer's theorem]\label{th:mercer}
    Let \(K\) be a continuous kernel on \([0,1]^2\). Then, there exists an orthonormal basis \(\{\phi_j\}\) of \(L^2[0,1]\) consisting of eigenfunctions of \(\mathcal K\), whose corresponding eigenvalues are all positive and form a decreasing sequence \(\{\lambda_j\}\to 0\), and such that
    \[
      K(t, s) = \sum_{j=1}^\infty \lambda_j \phi_j(t)\phi_j(s)
    \]
    for all \(t\) and \(s\), where the convergence is absolute and uniform.
\end{theorem}

@@

\begin{theorem}[Karhunen-Loève expansion]
  With the assumptions and notations of Theorem~\eqref{th:mercer}, the centered process \(X=X(t)\) with covariance function \(K\) admits the quadratic-mean representation
  \[
  X(t) = \sum_{j=1}^\infty \zeta_j \phi_j(t),
  \]
  where the convergence is in \(L^2(\Omega)\) and uniform in \(t\). The \(\zeta_j\) are independent zero-mean random variables with \(\E[\zeta_i\zeta_j]=\delta_{ij}\lambda_j\), given explictly by the formula
  \[
    \zeta_j = \int_0^1 X(t)\phi_j(t)\, dt.
  \]
\end{theorem}

In addition to these results, it is worth mentioning that the covariance operator \(\mathcal K\) provides several alternative definitions of \(\Hcal(K)\). For example, the RKHS can be identified with the image of the operator's square root, i.e., \(\Hcal(K) = \mathcal K^{1/2}(L^2[0, 1])\), with inner product \(\dotprod{f}{g}_K = \dotprod{\mathcal K^{-1/2}(f)}{\mathcal K^{-1/2}(g)}\). Furthermore, we can think of the norm in \(\Hcal(K)\) as an \(L^2\)-like regularized norm, since this space can be seen as \(\Hcal(K) = \{f \in L^2[0, 1]: \ \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}^2 < \infty \}\), with corresponding inner product \(\dotprod{f}{g}_K = \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}\dotprod{g}{\phi_j}\). Note that, since \(\{\lambda_j\}\) tends to zero, this definition highlights the fact that functions in \(\Hcal(K)\) are smooth, not only in that they inherit the regularity of the kernel \(K\), but also in the sense that their components in an orthonormal basis, namely \(\dotprod{f}{\phi_j}\), need to vanish quickly.

@@

Until now we have seen that \(\Hcal(K)\) is a space related to \(X\) only through its covariance function. Nevertheless, we will now hopefully clarify that this space can indeed be seen as \textit{the} Hilbert space inherently associated with the process \(X\) itself. Let us shift the perspective momentarily and consider the linear span of \(X\) in the space of all zero-mean random variables with finite second moment, i.e.:
\[
\mathcal L_0(X) = \left\{\sum_{i=1}^p a_i X(t_i): p\in\N, \ a_i \in \R, \ t_i \in [0,1]\right\}.
\]
It is well known that its completion in \(L^2(\Omega)\), denoted \(\mathcal L(X)\), is a Hilbert space, sometimes called the Hilbert space generated by \(X\). In the words of \citet{berlinet2004reproducing}, ``\(\mathcal L(X)\) contains the random variables attainable by linear operations, including limits, on the measurements of the process''. As it happens, via \textit{Loève's isometry} \citep{loeve1948fonctions} one can establish a congruence \(\Psi_X\) between \(\Hcal(K)\) and \(\mathcal L(X)\) \citep[see Lemma 1.1 in][]{lukic2001stochastic}. This isometry is essentially the completion of the correspondence
  \begin{equation}\label{eq:loeves-isometry}
  \sum_{i=1}^p a_i X(t_i) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
  \end{equation}
and can be formally defined, in terms of its inverse, as \(\Psi^{-1}_X(U)(t) = \E[U X(t)]\) for \(U \in \mathcal L(X)\). Thus, \(\Hcal(K)\) can be regarded (and it is often done in statistics) as an isometric copy of \(\mathcal L(X)\).

However, despite the close connection between the process \(X\) and the space \(\Hcal(K)\), special care must be taken when dealing with concrete realizations of the process. It can be shown that under rather general conditions the trajectories of \(X\) \textit{do not belong} to the corresponding RKHS with probability one \citep[see for example][Cor.~7.1]{lukic2001stochastic}. For instance, the trajectories of the Brownian motion are known to be nowhere differentiable, but as shown in Example \ref{ex:bm}, the elements of the corresponding RKHS are differentiable almost everywhere. An heuristic argument to justify this fact, given in \citet{wahba1990spline}, is as follows: consider the Karhunen-Loève expansion of \(X\), i.e.,
\[
  X(t) = \sum_{j=1}^\infty \zeta_j \phi_j(t),
\]
and the truncated version \(X_N(t)\) up to the \(N\)-th term. On the one hand, for each fixed \(t\) we have \(X_N(t)\to X(t)\) in the quadratic mean sense (by the Karhunen-Loève theorem), but on the other hand observe that
\[
  \E\|X_N(\cdot)\|^2_K = \E \sum_{j=1}^N \frac{\zeta_j^2}{\lambda_j} = N \to \infty \quad (N\to\infty).
\]

@@

\begin{proposition}
  For every \(t\in[0,1]\) and \(f, g \in \Hcal(K)\), the following relations are satisfied:
\begin{enumerate}
  \item \(\dotprod{X}{K(t, \cdot)}_K = X(t)\), a particular reproducing property.
  \item \(\E[\dotprod{X}{f}_K] = 0\).
  \item \(\E[\dotprod{X}{f}_K\dotprod{X}{g}_K] = \dotprod{f}{g}_K\).
\end{enumerate}
\end{proposition}
Note that the statement \((iii)\) above provides yet another characterization of the inner product in \(\Hcal(K)\), where \(\dotprod{X}{f}_K\) and \(\dotprod{X}{g}_K\) are understood as the random variables representing \(f\) and \(g\) in \(\mathcal L(X)\), respectively.

@@

Although not the main concern of this work, the theory of reproducing kernels and the associated kernel methods find applications in many areas of machine learning. For example, the well-known \textit{kernel trick} can be seen as a specific application of the reproducing property, where a feature map \(\Phi(x) \in \mathcal H\) is applied to the objects in the space of interest, transforming them into richer elements in a RKHS. Then, many computations can be efficiently carried out in an implicit manner by means of the corresponding kernel
\[
K(x, y)=\dotprod{\Phi(x)}{\Phi(y)}_{\mathcal H}.
\]
The fact that one does not need to explicitly compute \(\Phi(x)\) is especially relevant for kernels with an simple expression but a complex feature space (e.g. infinite-dimensional), such as the Gaussian kernel \(K(x, y)=\exp(-\gamma\|x-y\|^2)\).

@@

\begin{theorem}[Generalized representer theorem]
  Consider a RKHS \(\Hcal(K)\) of functions defined over a non-empty set \(\mathcal X\), a training set \(\{(x_i, y_i)\}_{i=1}^p \subset (\mathcal X \times \R)^p\), an arbitrary error function \(R: (\mathcal X  \times \R^2)^p \to \R\cup\{\infty\}\), and a strictly increasing real-valued function \(\Omega:[0,\infty)\to \R\). Then, any minimizer \(f^\ast\) of the regularized empirical risk
  \[
    R(\{(x_i, y_i, f(x_i))\}_{i=1}^p) + \Omega(\|f\|_K)
  \]
  is of the form \(f^\ast(\cdot) = \sum_{i=1}^p a_i K(x_i, \cdot)\), where \(a_i\in\R\) for all \(i=1,\dots,p\).
\end{theorem}

@@

When \(x\in\R\), we can use Monte Carlo algorithms such as \textit{rejection sampling}. In a nutshell, this method samples uniformly from the area under \(f(x)\) by means of an auxiliary density that encompasses this area. Specifically, one needs to find a density function \(g\) with \(\supp f \subset \supp g\) and \(f \leq Mg\) for some \(M>1\). Then, sampling from \(g\) and accepting the proposal, say \(x\), with probability \(f(x)/Mg(x)\) yields an approximate set of samples from \(f\). Nevertheless, these kinds of algorithms are rapidly affected by the so-called \textit{curse of dimensionality}, so new techniques are needed when sampling from multidimensional distributions. This is where Markov chain Monte Carlo (MCMC) methods come into play \citep[see e.g.][]{brooks2011handbook}.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.25\textwidth]{cube_sphere}
  \caption{To cover the volume of \(f(x)\) in higher dimensions we intuitively need a volume that grows exponentially with the dimension itself.}
\end{figure}

MCMC methods are based on the iterative construction of a Markov chain whose stationary distribution is the objective distribution \(\pi(x)\). In this way, we can take the samples of a sufficiently advanced chain as approximate realizations of \(X\sim \pi(x)\). When it comes to building the chains, transition probabilities are assigned in such a way that regions with higher density with respect to \(\pi(x)\) are favored (by means of the known proportional function \(f(x)\)). The dimensionality issues are somewhat mitigated, and on top of that there are specific tuning techniques to tackle them. However, the samples obtained from this procedure are not independent, though they can be made so by \textit{thinning} the chain and considering, say, only one every \(M\) samples.

>>>>

 The most common scalar-on-function linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the monograph by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^p\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is a random error term independent from \(X\) with \(\E [\epsilon]=0\) and \(\E[X(t)\epsilon]=0\) almost everwhere; and the functional slope parameter \(\beta=\beta(\cdot)\) is assumed to be a member of the infinite-dimensional space \(L^2[0, 1]\). A careful rearrangement of~\eqref{eq:l2-linear-model} shows that the model is equivalently expressed as \(\Delta=\K\beta\), where \(\Delta(\cdot)\) is the cross-covariance function of \(X\) and \(Y\), and \(\K\) is the covariance operator of the process \(X\) (to be defined in Section~\ref{sec:rkhs}). This expression is a continuous analog of the normal equations that arise in finite-dimensional regression settings, and since the operator \(\K\) is non-invertible in general, questions of existence and unicity of \(\beta\) need further study (see \citet{cardot2011functional} for more insights).

@@

A common inference strategy is to expand both \(X\) and \(\beta\) on a certain data-driven orthonormal basis of \(L^2[0,1]\), say \(\{\phi_j\}\), up to a certain value \(p\in\N\):
\[
  X(t)=\sum_{j=1}^p Z_j \phi_j(t), \quad \beta(t) = \sum_{j=1}^p \beta_j \phi_j(t).
\]
In this way, model~\eqref{eq:l2-linear-model} simplifies to
\[
Y = \alpha_0 + \sum_{j=1}^p \beta_j Z_j,
\]
which can then be solved in the usual manner. For example, when the chosen basis is composed of eigenfunctions of the covariance operator \(\K\), this method is known as Functional Principal Component Regression (FPCR). Another class of methods focus on selecting an \textit{a priori} basis for \(\beta\) (e.g. a spline basis) and solving a penalized least squares problem. In particular, for a truncated basis representation of the form \(\beta(t)=\phi(t)'b\), with \(\phi(t)=(\phi_1(t), \dots, \phi_p(t))'\) and \(b\in\R^p\), a typical minimization problem would be
\[
\argmin_{(\alpha_0, b)\in\R\times\R^p} \ \left\{ \sum_{i=1}^n \left( y_i - \alpha_0 - \int_0^1 x_i(t)[\phi(t)'b]\, dt\right)^2 + \lambda \Omega(b) \right\},
\]
where \(\Omega\) is some penalty function and \(\lambda>0\) is a regularization parameter.

Turning our attention to logistic regression, a similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][Sec.~3.2]{berrendero2018functional}.

@@

The term Bayesian inference usually refers to a wide class of methods that to some extent employ Bayes' theorem to update the initial probability assigned to a hypothesis when new information is available. It can be seen as a general tool for modeling situations or problems that involve uncertainty; some examples include Bayesian hierarchical modeling, Bayesian regression or even Bayesian neural networks \citep[see e.g.][]{murphy2012machine, bishop2006pattern}. Specifically, in this work we will be interested in performing parameter estimation in a Bayesian framework, so that pre-existing information and beliefs about the parameters can be incorporated into the model.

One difficulty found in almost all Bayesian methods is that the posterior distribution, the main object of interest, is usually intractable due to the integral that appears as the normalizing constant in Bayes' rule. A way to bypass this limitation is to use conjugate distributions, where it can be rigorously proven that a certain combination of likelihood and prior distributions produces a posterior distribution in the same family as the prior. However, unless conjugate priors are used, a closed-form expression is generally unattainable and some type of approximation is required. Apart from basic numerical integration, some well-performing methods in this regard include variational inference approaches \citep[e.g.][]{blei2017variational} and MCMC methods \citep[e.g.][]{brooks2011handbook}. The former are based on approximating the posterior by another distribution restricted to a certain parametric family so that the Kullback-Leibler divergence between them is minimized, whereas the latter are iterative methods that directly provide approximate samples of the posterior distribution.

As a comment, there are some works that tackle Bayesian inference from a functional perspective; a recent example are the functional Bayesian neural networks proposed by~\citet{sun2019functional}.

@@

In this section we present a brief exposition of some basic concepts regarding reproducing kernel Hilbert spaces. Although there are quite a few ways of introducing these spaces, we adopt a probabilistic point of view that will set the stage for the development of the subsequent theory. For a more detailed account, a good reference is the book by~\citet{berlinet2004reproducing}.

@@

\begin{definition}
  We say that a function of two variables \(K: \T \times \T \to \R\) is positive semidefinite\footnote{Sometimes in the literature these functions are known simply as positive definite.} if
  \[
    \sum_{i,j=1}^p a_ia_jK(t_i, t_j) \geq 0
  \]
  for any \(p\in\N\), any \((a_1,\dots, a_p)\in \R^p\) and any \((t_1,\dots, t_p)\in\T^p\). Note that this is equivalent to saying that the matrix \((K(t_i, t_j))_{i,j}\) is positive semidefinite for any choice of \(p\in\N\) and \((t_1,\dots, t_p)\in\T^p\).
\end{definition}

@@

\begin{proposition} The following properties hold for the space \(\Hcal(K)\):
  \begin{enumerate}
    \item The evaluation operator \(\delta_t\) is bounded for all \(t\in[0,1]\).
    \item Norm convergence implies pointwise convergence, and if \(K\) is continuous, it also implies uniform convergence.
    \item If \(K\) is m-times continuously differentiable, then so is every function \(f\in\Hcal(K)\). In particular, if \(K\) is continuous, every function \(f\in\Hcal(K)\) is continuous.
  \end{enumerate}
\end{proposition}
\begin{proof}
  To prove \((i)\), simply note that for any \(t\in[0,1]\) and \(f\in\Hcal(K)\) the Cauchy-Schwarz inequality and the reproducing property tell us that
  \[
  |\delta_t(f)| = |f(t)| = |\dotprod{f}{K(t, \cdot)}_K| \leq \|K(t, \cdot)\|_K \|f\|_K = \sqrt{K(t, t)}\|f\|_K,
  \]
  and consequently \(\|\delta_t\|\leq \sqrt{K(t, t)}\). To see \((ii)\), observe that for any \(t\in[0,1]\) we have
  \[
    |f_n(t) - f(t)| = |\delta_t(f_n - f)| \leq \|\delta_t\|\|f_n - f\|_K,
  \]
  and we have just shown above that the evaluation operator is bounded. Moreover, if \(K\) is continuous on its (compact) domain, \(\|\delta_t\| \leq \sup_s \sqrt{K(s, s)}=M<\infty\) for all \(t\), so the convergence is indeed uniform. Finally, we prove the second statement in \((iii)\), and refer the reader to \citet[][Th 2.6]{saitoh2016theory} for a complete proof. If \(K\) is continuous, then it is clear that every function in \(\Hcal_0(K)\) is continuous. Let us now fix an arbitrary \(f\in \Hcal(K)\). By definition, there exists a sequence \(\{f_n\}\subset \Hcal_0(K)\) of continuous functions such that \(\|f_n - f\|_K \to 0\). But then \(f_n \to f\) uniformly by \((ii)\), and therefore \(f\) is continuous as the uniform limit of continuous functions. Note that this statement remains valid, with a slightly different proof, even when the underlying domain is not compact.
\end{proof}

@@

\begin{example}[Finite-dimensional RKHS]
  Setting aside our stochastic process framework for a moment, we can conceive examples outside of \(\T =[0,1]\). Consider a vector-valued random variable \(X = (X_1, \dots, X_p)\) with non-singular covariance matrix \(\Sigma\). In this case the index set would be \(\T_p = \{1,\dots,p\}\), so identifying functions on \(\T_p\) with points in the \(p\)-dimensional Euclidean space, we have \(\Hcal(\Sigma)=\R^p\), with inner product \(\dotprod{x}{y}_\Sigma=x\Sigma^{-1}y\).
\end{example}

@@

 Note that, as expected, the covariance function of \(X\) plays a crucial role in characterizing the associated RKHS. An integral operator closely related to the kernel is the so-called covariance operator, namely
\[
\K f(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f \in L^2[0, 1],
\]
which is self-adjoint and compact when \(K\) is continuous \citep[e.g.][Th.~4.6.2]{hsing2015theoretical}, so for the remaining of this work we will indeed suppose that \(K\) is continuous. As it turns out, the covariance function admits a spectral decomposition in terms of the eigenvalues and eigenfunctions of this integral operator, a sort of continuous generalization of the eigendecomposition of a symmetric positive semidefinite matrix.

\begin{theorem}[Mercer's theorem]\label{th:mercer}
    Let \(K\) be a continuous kernel on \([0,1]^2\). Then, there exists an orthonormal basis \(\{\phi_j\}\) of \(L^2[0,1]\) consisting of eigenfunctions of \(\K\), whose corresponding eigenvalues are all positive and form a decreasing sequence \(\{\lambda_j\}\to 0\), and such that
    \[
      K(t, s) = \sum_{j=1}^\infty \lambda_j \phi_j(t)\phi_j(s)
    \]
    for all \(t\) and \(s\), where the convergence is absolute and uniform.
\end{theorem}

@@

\begin{theorem}[Karhunen-Loève expansion]
  With the assumptions and notations of Theorem~\eqref{th:mercer}, the centered process \(X=X(t)\) with covariance function \(K\) admits the quadratic-mean representation
  \[
  X(t) = \sum_{j=1}^\infty \zeta_j \phi_j(t),
  \]
  where the convergence is in \(L^2(\Omega)\) and uniform in \(t\). The \(\zeta_j\) are independent zero-mean random variables with \(\E[\zeta_i\zeta_j]=\delta_{ij}\lambda_j\), given explictly by the formula
  \[
    \zeta_j = \int_0^1 X(t)\phi_j(t)\, dt, \quad j\geq 1.
  \]
\end{theorem}

In addition to these results, it is worth mentioning that the covariance operator \(\K\) provides several alternative definitions of \(\Hcal(K)\). For example, the RKHS can be identified with the image of the operator's square root, i.e., \(\Hcal(K) = \K^{1/2}(L^2[0, 1])\), with inner product \(\dotprod{f}{g}_K = \dotprod{\K^{-1/2}(f)}{\K^{-1/2}(g)}\). Furthermore, we can think of the norm in \(\Hcal(K)\) as an \(L^2\)-like regularized norm, since this space can be seen as \(\Hcal(K) = \{f \in L^2[0, 1]: \ \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}^2 < \infty \}\), with corresponding inner product \(\dotprod{f}{g}_K = \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}\dotprod{g}{\phi_j}\). Note that, since \(\{\lambda_j\}\) tends to zero, this definition highlights the fact that functions in \(\Hcal(K)\) are smooth, not only in that they inherit the regularity of the kernel \(K\), but also in the sense that their components in an orthonormal basis, namely \(\dotprod{f}{\phi_j}\), need to vanish quickly.

@@

Until now we have seen that \(\Hcal(K)\) is a space related to \(X\) only through its covariance function. Nevertheless, we will now hopefully clarify that this space can indeed be seen as \textit{the} Hilbert space inherently associated with the process \(X\) itself. Let us shift the perspective momentarily and consider the linear span of \(X\) in the space of all zero-mean random variables with finite second moment, i.e.:
\[
\Lcal_0(X) = \left\{\sum_{i=1}^p a_i X(t_i): p\in\N, \ a_i \in \R, \ t_i \in [0,1]\right\}.
\]
It is well known that its completion in \(L^2(\Omega)\), denoted \(\Lcal(X)\), is a Hilbert space, sometimes called the Hilbert space generated by \(X\). In the words of \citet{berlinet2004reproducing}, ``\(\Lcal(X)\) \textit{contains the random variables attainable by linear operations, including limits, on the measurements of the process}''. As it happens, via \textit{Loève's isometry} \citep{loeve1948fonctions} one can establish a congruence \(\Psi_X\) between \(\Hcal(K)\) and \(\Lcal(X)\) \citep[see Lemma 1.1 in][]{lukic2001stochastic}. This isometry is essentially the completion of the correspondence
  \begin{equation}\label{eq:loeves-isometry}
  \sum_{i=1}^p a_i X(t_i) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
  \end{equation}
and can be formally defined, in terms of its inverse, as \(\Psi^{-1}_X(U)(t) = \E[U X(t)]\) for \(U \in \Lcal(X)\). Thus, \(\Hcal(K)\) can be regarded as an isometric copy of \(\Lcal(X)\), an approach that is often useful in statistics.

However, despite the close connection between the process \(X\) and the space \(\Hcal(K)\), special care must be taken when dealing with concrete realizations of the process. It can be shown that under rather general conditions the trajectories of \(X\) \textit{do not belong} to the corresponding RKHS with probability one \citep[see for example][Cor.~7.1]{lukic2001stochastic}. For instance, the trajectories of a Brownian motion are known to be nowhere differentiable, but as shown in Example~\ref{ex:bm}, the elements of the corresponding RKHS are differentiable almost everywhere. An heuristic argument to justify this fact, given in \citet{wahba1990spline}, is as follows: consider the Karhunen-Loève expansion of \(X\), i.e.,
\[
  X(t) = \sum_{j=1}^\infty \zeta_j \phi_j(t),
\]
and the truncated version \(X_N(t)\) up to the \(N\)-th term. On the one hand, for each fixed \(t\) we have \(X_N(t)\to X(t)\) in the quadratic mean sense (by the Karhunen-Loève theorem), but on the other hand observe that
\[
  \E\|X_N(\cdot)\|^2_K = \E \sum_{j=1}^N \frac{\zeta_j^2}{\lambda_j} = N \to \infty \quad (N\to\infty).
\]

@@

\begin{proposition}
  For every \(t\in[0,1]\) and \(f, g \in \Hcal(K)\), the following relations are satisfied:
\begin{enumerate}
  \item \(\dotprod{X}{K(t, \cdot)}_K = X(t)\), a particular reproducing property.
  \item \(\E[\dotprod{X}{f}_K] = 0\).
  \item \(\E[\dotprod{X}{f}_K\dotprod{X}{g}_K] = \dotprod{f}{g}_K\).
\end{enumerate}
\end{proposition}
Note that the statement \((iii)\) above provides yet another characterization of the inner product in \(\Hcal(K)\), where \(\dotprod{X}{f}_K\) and \(\dotprod{X}{g}_K\) are understood as the random variables representing \(f\) and \(g\) in \(\Lcal(X)\), respectively.

@@

Although not the main concern of this work, the theory of reproducing kernels and the associated kernel methods find applications in many areas of machine learning. For example, the well-known \textit{kernel trick} can be seen as a specific application of the reproducing property, where a feature map \(\Phi(x) \in \Hcal\) is applied to the objects in the space of interest, transforming them into richer elements in a RKHS. Then, many computations can be efficiently carried out in an implicit manner by means of the corresponding kernel
\[
K(x, y)=\dotprod{\Phi(x)}{\Phi(y)}_{\Hcal}.
\]
The fact that one does not need to explicitly compute \(\Phi(x)\) is especially relevant for kernels with an simple expression but a complex feature space (e.g. infinite-dimensional), such as the Gaussian kernel \(K(x, y)=\exp(-\gamma\|x-y\|^2)\).

@@

\begin{theorem}[Generalized representer theorem]
  Consider a RKHS \(\Hcal(K)\) of functions defined over a non-empty set \(\X\), a training set \(\{(x_i, y_i)\}_{i=1}^n \subset (\X \times \R)^n\), an arbitrary error function \(R: (\X  \times \R^2)^n \to \R\cup\{\infty\}\), and a strictly increasing real-valued function \(\Omega:[0,\infty)\to \R\). Then, any minimizer \(f^\ast\) of the regularized empirical risk
  \[
    R(\{(x_i, y_i, f(x_i))\}_{i=1}^n) + \Omega(\|f\|_K)
  \]
  is of the form \(f^\ast(\cdot) = \sum_{i=1}^n a_i K(x_i, \cdot)\), where \(a_i\in\R\) for all \(i=1,\dots,n\).
\end{theorem}

@@

When \(x\in\R\), we can use Monte Carlo algorithms such as \textit{rejection sampling}. In a nutshell, this method samples uniformly from the area under \(f(x)\) by means of an auxiliary density that encompasses this area. Specifically, one needs to find a density function \(g\) with \(\supp f \subset \supp g\) and \(f \leq Mg\) for some \(M>1\). Then, sampling from \(g\) and accepting the proposal, say \(x\), with probability \(f(x)/Mg(x)\) yields an approximate set of samples from \(f\). Nevertheless, these kinds of algorithms are rapidly affected by the so-called \textit{curse of dimensionality} (see Figure \ref{fig:curse_dimensionality}), so new techniques are needed when sampling from multidimensional distributions. This is where Markov chain Monte Carlo (MCMC) methods come into play.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.25\textwidth]{cube_sphere}
  \caption{To cover the volume of \(f(x)\) in higher dimensions we intuitively need a volume that grows exponentially with the dimension itself.}\label{fig:curse_dimensionality}
\end{figure}

MCMC methods are based on the iterative construction of a Markov chain whose stationary distribution is the objective distribution \(\pi(x)\). In this way, we can take the samples of a sufficiently advanced chain as approximate realizations of \(X\sim \pi(x)\). When it comes to building the chains, transition probabilities are assigned in such a way that regions with higher density with respect to \(\pi(x)\) are favored (by means of the known proportional function \(f(x)\)). The dimensionality issues are somewhat mitigated, and on top of that there are specific tuning techniques to tackle them. However, the samples obtained from this procedure are not independent, though they can be made so by \textit{thinning} the chain and considering only one every few samples.
